---
title: "Human Activity Tracking Codebook"
author: "Dominik Sudwischer"
date: "22 October 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Extracting the Required Base Data

The original data set comprises 10299 obvervations collected from 30 volunteers that participated in a study capturing different movement measurements via the sensors of a Samsung Galacy S2 smartphone. According to the description of the original data set, the following facts hold true: The six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) had their corresponding motions tracked by an embedded accelerometer and a gyroscope. To be more precise, 3-axial linear acceleration and angular velocity have been captured at a constant rate of 50 measurements per seconds. The labeling of the data has been performend manually using recorded video files. In the original data, the 10299 observations have been split into two datasets with 70% ('X_train.txt') and 30% ('X_test.txt') of the records respectively. Originally, there are 561 features. The corresponding files 'y_train' and 'y_test' contain the id of the subject. The text file 'activity labels.txt' explain each activity id, a number between 1 and 6, in plain text.

These 5 files were concatenated in the following way: for $s \in \{\rm{"train"}, \rm{"test"}\}$, X_s and y_s have been combined using cbind. After that, both data frames have been concatenated using rbind. The activity ids have been replaced the the actual name strings for each activity that can be seen in 'activity labels.txt'. 

## Changes to Variable Names

The names have been changed to be more readable by humans. In particular, leading 't' and 'f' have been renamed to 'time' and 'frequency', respectively. Also, 'Acc', 'Gyro' and 'Mag' have been renamed to 'acceleration', 'gyroscope' and 'magnitude'. All names have been transformed to lower case.

## Included Variables

After the initial transformations performed by the run_analysis script the data consists of a single large data frame including a total of 10299 records and 68 variables, 66 of which are extracted from the original dataset. The last two ones are the subject ID, an integer between 1 and 30, and the activity as a captial letter string. The 66 extracted variables are those that are mean and standard deviations of measurements. Please note that frequency means ('MeanFreq()') are not included.

The resulting data set comprises means of each of those 66 variables, grouping the records by subject id and activity. The following variables are contained:

```{r feature_names, echo = FALSE}
features <- readLines("feature_names.txt")
features <- gsub("[\\d]+ ", "", features)
features
```

Each of these values have been generated by grouping the original data set by subject and activity and then computing the mean for each of the resulting 180 groups. That is, it contains the mean value of every single variable for each of the 30 subjects doing each of the six activities.

## Units 

The units are the same as in the original data. Please note that each feature has been normalized in the original data to be in [-1, 1]. The exact methods of generating the data, including filtering noise and applying Fourier transformations, can be looked up in the documentation of the original dataset. More detailed information can be found at http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones.

## Original CodeBook
==================================================================
Human Activity Recognition Using Smartphones Dataset
Version 1.0
==================================================================
Jorge L. Reyes-Ortiz, Davide Anguita, Alessandro Ghio, Luca Oneto.
Smartlab - Non Linear Complex Systems Laboratory
DITEN - Università degli Studi di Genova.
Via Opera Pia 11A, I-16145, Genoa, Italy.
activityrecognition@smartlab.ws
www.smartlab.ws
==================================================================

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:
======================================

- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

The dataset includes the following files:
=========================================

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The following files are available for the train and test data. Their descriptions are equivalent. 

- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

- 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

- 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

- 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

Notes: 
======
- Features are normalized and bounded within [-1,1].
- Each feature vector is a row on the text file.

For more information about this dataset contact: activityrecognition@smartlab.ws

License:
========
Use of this dataset in publications must be acknowledged by referencing the following publication [1] 

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.